---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# spatialDE

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R-CMD-check-bioc](https://github.com/sales-lab/spatialDE/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/sales-lab/spatialDE/actions)
[![Codecov test coverage](https://codecov.io/gh/sales-lab/spatialDE/branch/main/graph/badge.svg)](https://codecov.io/gh/sales-lab/spatialDE?branch=main)
[![BioC release status](http://www.bioconductor.org/shields/build/release/bioc/spatialDE.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/spatialDE)
[![BioC devel status](http://www.bioconductor.org/shields/build/devel/bioc/spatialDE.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/spatialDE)
<!-- badges: end -->

The **spatialDE** package provides an R wrapper for the Python SpatialDE library,
using `r BiocStyle::CRANpkg("reticulate")` and `r BiocStyle::Biocpkg("basilisk")`.

[SpatialDE](https://github.com/Teichlab/SpatialDE), by [Svensson et al., 2018][Svensson2018],
is a method to identify spatially variable genes (SVGs) in spatially resolved transcriptomics data.

This package started as part of the
[BiocSpatialChallenges](https://helenalc.github.io/BiocSpatialChallenges/index.html).


## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then
install `r BiocStyle::Biocpkg("spatialDE")` from
[*Bioconductor*](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("spatialDE")
```

The development version of **spatialDE** can be installed from
[GitHub](https://github.com/sales-lab/spatialDE) with:

```{r 'install_dev', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
    
BiocManager::install("sales-lab/spatialDE")
```


## Basic usage

```{r usage, eval=TRUE}
library(spatialDE)
spe <- mockSVG(return_SPE = TRUE)
de_results <- spatialDE(spe)
head(de_results)
```


## Citation

Below is the citation output from using `citation('spatialDE')` in R. Please
run this yourself to check for any updates on how to cite __spatialDE__.

Please note that this package merely provides a wrapper to use the original Python methods in R. If you find these methods useful, please also consider citing the [original paper][Svensson2018].

```{r 'citation', eval = requireNamespace('spatialDE'), echo = FALSE, comment = ""}
print(citation('spatialDE'), bibtex = TRUE)
```


## Code of Conduct

Please note that the **spatialDE** project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.


## Useful links

- `r BiocStyle::Biocpkg("SpatialExperiment")`
- [BiocSpatialChallenges](https://helenalc.github.io/BiocSpatialChallenges/index.html)

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.

[Svensson2018]: https://doi.org/10.1038/nmeth.4636
