---
title: "Introduction to spatialDE"
author: 
  - name: Davide Corso
    affiliation:
    - University of Padova
    email: davide.corso.2@phd.unipd.it
  - name: Milan Malfait
    affiliation:
    - Ghent University
    email: milan.malfait94@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
    df_print: paged
date: "`r BiocStyle::doc_date()`"
vignette: >
  %\VignetteIndexEntry{Introduction to spatialDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

[SpatialDE](https://github.com/Teichlab/SpatialDE) by [Svensson et al., 2018](https://www.nature.com/articles/nmeth.4636), is a method to identify spatially variable genes (SVGs) in spatially resolved transcriptomics data.


# Installation

```{r, eval=FALSE}
## Install development version from GitHub
BiocManager::install("sales-lab/spatialDE")
```


# Example: [Mouse Olfactory Bulb](https://github.com/Teichlab/SpatialDE/tree/master/Analysis/MouseOB)

Reproducing the
[example analysis](https://github.com/Teichlab/SpatialDE#spatialde-significance-test-example-use) from the original `r basilisk::PyPiLink("SpatialDE")` Python package.

```{r setup}
library(spatialDE)
library(ggplot2)
```


## Load data

```{r}
# Expression file used in python SpatialDE
# 8.4 Mb
expression_csv <-
  "https://media.githubusercontent.com/media/Teichlab/SpatialDE/ad411d51e887575a34c90d3f12fd9c6844e35c9f/Analysis/MouseOB/data/Rep11_MOB_0.csv"

# Sample Info file used in python SpatialDE
# 7 KB
coordinate_csv <-
  "https://media.githubusercontent.com/media/Teichlab/SpatialDE/master/Analysis/MouseOB/MOB_sample_info.csv"
```

Loading the `expression_csv` file.
Note that this data has __genes in columns and samples in rows__.

```{r}
df <- read.csv(expression_csv, header = TRUE, row.names = 1)
df[1:5, 1:5]
```

Loading the `coordinate_csv` file.

```{r}
sample_info <- read.csv(coordinate_csv, row.names = 1)
head(sample_info)
```

### Filter out pratically unobserved genes

```{r}
df <- df[, colSums(df) >= 3]
```

### Get total_counts for every spot

```{r}
df <- df[row.names(sample_info), ]
sample_info$total_counts <- rowSums(df[, ])
head(sample_info)
```

### Get coordinates from `sample_info`

```{r}
X <- sample_info[, c("x", "y")]
head(X)
```

## `stabilize`

The SpatialDE method assumes normally distributed data, so we stabilize the variance of the negative binomial distibuted counts data using Anscombe's approximation.
The `stabilize` function takes as input a `data.frame` of expression values with samples in columns and genes in rows. Thus, in this case, we have to transpose the data.

```{r stabilize}
tdf <- t(df)

norm_expr <- spatialDE::stabilize(tdf)
as.data.frame(norm_expr)[1:5, 1:5]
```

## `regress_out`

Next, we account for differences in library size between the samples by regressing out the effect of the total counts for each gene using linear regression.

```{r regres_out}
resid_expr <- regress_out(sample_info, norm_expr)
as.data.frame(resid_expr)[1:5, 1:5]
```

## `run`

To reduce running time, the SpatialDE test is run on a subset of 1000 genes.
Running it on the complete data set takes about 10 minutes.

```{r run-spatialDE}
# results <- spatialDE::run(X, resid_expr)

# For this example, run spatialDE on the first 1000 genes
sample_resid_expr <- head(resid_expr, 1000)

results <- spatialDE::run(X, sample_resid_expr)
head(results[order(results$qval), ])
```

## `model_search`

Finally, we can classify the DE genes to interpetable DE classes using the `model_search` function.
We apply the model search on filtered DE results, using a threshold of 0.05 for the Q-values.

```{r model_search}
de_results <- results[results$qval < 0.05, ]
# ms_results <- spatialDE::model_search(X, resid_expr, de_results)
ms_results <- spatialDE::model_search(X, sample_resid_expr, de_results)
head(ms_results[order(ms_results$qval), ])
```

## `spatial_patterns`

Furthermore, we can group spatially variable genes (SVGs) into spatial patterns using automatic expression histology (AEH).

```{r spatial_patterns}
sres <- results[results$qval < 0.05, ]

sp <- spatial_patterns(X, sample_resid_expr, sres, 4L, 1.5)
sp$pattern_results
```

## Plots

Visualizing one of the most significant genes.

```{r}
gene <- "Pcp4"

g <- ggplot(data = sample_info, aes(y = y, x = x, color = norm_expr[gene, ])) +
    geom_point(size = 7) +
    ggtitle(gene)

g1 <- g + scale_color_viridis_c() + labs(color = gene)

g1
```

# `sessionInfo` {-}

<details><summary>Session info</summary>

```{r session_info, echo=FALSE, cache=FALSE}
Sys.time()
sessionInfo()
```

</details>
